A Reimplementation of Python's dict using Simple Tabulation Hashing and Linear Probing

### 2013-04-22 docs/paper_lncs.tex
New conclusion

\subsection{Implementation Details}
We modified the source code of CPython 2.7.3 to use simple tabulation hashing
with linear probing.  We use a fixed number of tabulation tables and cycle between them
for variable length strings.

@@

We also experimented with indexing using every pair of bytes in the string,
treating them as 16bit integers and
using $2^{16}$ tabulation table entries.

@@

After implementation and profiling, we found that dictionary table lookups had
over twice as many level 1 cache misses with tabulation hashing than without.
Hypothesizing that the tabulation tables were evicting other, more relevant
data from the level 1 cache.  To fix this, we inserted x86 memory prefetch
instructions into our hash code to tell the CPU's caching system that it should
deprioritize storing the tabulation tables in cache.  We used GCC's
\begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt} \cite{prefetch}
to generate these instructions.  However, this only yielded about a 1\% performance improvement,
negligible compared to the performance improvement of different hash schemes.

@@

\section{Conclusion and future work}
Throughout the course of our work we came to the following conclusions:
\begin{enumerate}
\item Hardware specific performance issues dominate at small scales: When you
are attempting to work on a process that runs as fast as a hash table lookup or
insertion already does, there is little room for error.  Improving hash
performance with simple tabulation is not an easy task simply because any
change you make will probably end up making the hash function run slower.  That said:

\item Simple tabulation yields slightly faster and more consistent performance:
With a careful implementation simple tabulation can yield increased hash
performance on linear probing work loads. Simple tabulation definitely leads to
shorter chain lengths, however, those shorter chain lengths do not always
justify the performance cost in all cases.

\item Memory latency presents serious but workable issues.  Doing a look up in the simple-tabulation table \emph{requires} a random memory access, exactly what we were trying to avoid. With careful memory management we can maintain good performance, but a solution requires careful thought about cache sizes and special purpose compiler directives.

\end{enumerate}

In terms of future work, we have 2 avenues of attack. First, we could use
different dictionaries for different purposes. Many of the Python dictionaries
are small, and their keysets never change. New built in functions are not added
during the runtime of the program and thus they could use perfect hashing.

@@

Given the strong performance of simple tabulation hashing and the only downside
being the cost of memory lookups, processor manufacturers should look into
adding random tables into hardware as a ROM.  We found that for what we found
optimal for Python, we only needed 16KB of randomness.  This would be a trivial
cost in ROM given that processors have MBs of higher level caches nowadays.

\bibliography{references}{}
\bibliographystyle{plain}

\appendix
\newpage
\section*{Appendix: Tornado Application}
Sample Tornado 2.2.1 application used for benchmark:
 \begin{verbatim}
import tornado.httpserver
import tornado.ioloop
import tornado.wsgi

def application(environ, start_response):
    status = '200 OK'
    output = 'Hello World!' 
    response_headers = [('Content-type', 'text/plain'),
        ('Content-Length', str(len(output)))]
    start_response(status, response_headers)
    return [output]
def main():
    container = tornado.wsgi.WSGIContainer(application)
    http_server = tornado.httpserver.HTTPServer(container)
    http_server.listen(8000)
    tornado.ioloop.IOLoop.instance().start()
if __name__ == "__main__":
    main()
 \end{verbatim}
\appendix

>>>>

\subsection{Implementation Details}
We modified the source code of CPython 2.7.3 to use simple tabulation hashing
with linear probing.  We use a fixed number of tabulation tables and cycle between them
for variable length strings. This does not provide the same level of theoretical guarantees for strings with length longer than the number of tables we have but empirically provides good results.

@@

We also experimented with indexing using every pair of bytes in the string,
treating them as 16bit integers and
using $2^{16}$ tabulation table entries but results were worse due to increased memory thrashing.

@@

After implementation and profiling, we found that dictionary table lookups had
over twice as many level 1 cache misses with tabulation hashing than without.
Hypothesizing that the cache misses were due to the tabulation tables not being in processor cache (16KB tables, 32KB L1 data, 3MB L2), we inserted x86 memory prefetch
instructions into our hash code to tell the CPU's caching system that it should
prioritize storing the tabulation tables in cache.  We used GCC's
\begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt} \cite{prefetch}
to generate these instructions.  However, this only yielded about a 1\% performance improvement in synthetic benchmarks,
negligible compared to the performance improvement of different hash schemes.

@@

\section{Conclusion and future work}
%Throughout the course of our work we came to the following conclusions:
%\begin{enumerate}
%\item Hardware specific performance issues dominate at small scales: When you
%are attempting to work on a process that runs as fast as a hash table lookup or
%insertion already does, there is little room for error.  Improving hash
%performance with simple tabulation is not an easy task simply because any
%change you make will probably end up making the hash function run slower.  That said:

Simple tabulation yields faster and more consistent hash performance than Python's current implementation  despite the increased number of memory accesses needed to compute the simple tabulation hash.  Our results show that performance improvements of up to 19\% are realizable in certain benchmarks with heavy lookup workloads such as class attribute lookups.  Overall improvement real python code performance improved by about 4\%.
%Simple tabulation definitely leads to shorter chain lengths, however, those shorter chain lengths do not %always justify the performance cost in all cases.

%Memory latency presents the most serious perofmrance issue.  Doing a look up in the simple-tabulation %table \emph{requires} a random memory access, exactly what we were trying to avoid. With careful memory %management we can maintain good performance, but a solution requires careful thought about cache sizes %and special purpose compiler directives.

In terms of future work, we have 2 avenues of attack. First, we could use
different dictionaries for different purposes. Many of the Python dictionaries
are small, and their keysets never change. New built in functions are not added
during the runtime of the program and thus they could use perfect hashing to improve lookups.

@@

Given the strong performance of simple tabulation hashing and the only downside
being the cost of memory lookups hampering performance, processor manufacturers should look into
adding random tables into hardware as a ROM.  We found that we only needed 16KB of randomness to acheive our 4\% overall performance gains for Python. This would be a trivial
cost in ROM given that processors have MBs of higher level caches nowadays.  An x86 instruction set extension providing access to random tables would benefit numerous applications that use dictionaries as a core data structures.

Overall, we suggest that the Python community should consider simple tabulation hashing with linear probing as a better implementation of Python's \textbf{dict} due to realizable performance improvements of at least 4\% and lessened performance variation.  In addition, if the tables are randomized at interpreter startup, the potential for malicious attacks against dictionary performance could be remedied.
\bibliography{references}{}
\bibliographystyle{plain}
